A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-02-06 analysis/paper/paper.Rmd
update paper

Most methods fulfilling most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`). Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`). However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding studies pinpointing these as good practices [@wilson_best_2014].

@@

To classify the output of methods and the gold standards, we defined a set of basic trajectory types. These trajectory types start from the most general trajectory type, a disconnected simple graph, and moves down in a directed acyclic graph using certain graph properties, progressively becoming more simple until the two basic types: linear and cycles `r ref("sfig", "trajectory_type_tree", "a")`. For every type, a corresponding directed graph type can be defined by allowing directedness `r ref("sfig", "trajectory_type_tree", "b")`.

@@

<p class='wip'> We wrapped each method (available at https://www.github.com/dynverse/dynmethods) and postprocessed its output into a common probabilistic trajectory model. This model consists of two parts. The milestone network represents the overall network topology, and contains of edges between different milestones and the length of the edge between them. The milestone percentages contain for each cell its position between two or more milestones, and sums for each cell to one. Linear methods are the easiest to wrap into this format, with two milestones and all cells in between. Methods such as 
</p>## Method quality control

@@

```{r}
add_sfig({
readRDS(figure_file("score_aspects.rds", experiment_id="score_rules/score_aspects"))
}, 
"score_aspects", 
"Different scores assess different aspects of the ", 
15, 
5
)
```

>>>>

Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`). Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`). However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding studies pinpointing these as good practices [@wilson_best_2014].

@@

To classify the output of methods and the gold standards, we defined a set of basic trajectory types. These trajectory types start from the most general trajectory type, a disconnected simple graph, and moves down in a directed acyclic graph using certain graph properties, progressively becoming more simple until the two basic types: linear and cycles (`r ref("sfig", "trajectory_type_tree", "a")`). For every type, a corresponding directed graph type can be defined by allowing directedness (`r ref("sfig", "trajectory_type_tree", "b")`).

@@

We wrapped each method (available at https://www.github.com/dynverse/dynmethods) and postprocessed its heterogenous output into a common probabilistic trajectory model. This model consists of two parts. The milestone network represents the overall network topology, and contains of edges between different milestones and the length of the edge between them. The milestone percentages contain for each cell its position between two or more milestones, and sums for each cell to one.

Depending on the output of a method, we used different strategies to convert the output to our common model:

* Linear trajectory methods usually return a single pseudotime value, which can be converted to our model using a single edge between two milestones
* Conversely, methods which assign each cell to a branch together with a pseudotime value and a branch network easily converted into our model
* Methods returning both a global pseudotime value and a probability for every end state were converted into our model by using one single begin milestone and an end milestone for every end state. The global pseudotime then determines the distance from the begin milestone, the rest of the cell's position is calculated by distributing the residual percentage over the end states, proportionally to the end state probabilities
Special conversions were necessary for certain method
* DPT: 
* Sincell
* SLICER
* SLICE
*

## Method quality control

@@

```{r}
add_sfig({
readRDS(figure_file("score_aspects.rds", experiment_id="score_rules/score_aspects"))
}, 
"score_aspects", 
"Different scores assess different aspects of the ", 
15, 
8
)
```
